---
title: "Data Cleaning with R"
author: "Josh Gardner"
date: "November 6, 2016"
output: html_document
---
#Install packages used in this session
```{r}
install.packages("dplyr", "stringr", "tidyr", "magrittr", "Hmisc", "ggplot2")
```

#Load data from CSV
```{r}
# make sure you set your working directory to the correct location before running this!

# load data; this data was downloaded from http://www.fec.gov/disclosurep/PDownload.do
infile = "P00000001D-ALL.csv"
cf = read.csv(infile, row.names = NULL, na.strings = c('', ' '))
```

#Inspect data
```{r}
head(cf, 20)

str(cf)

# view table of number of entries for each candidate 
table(cf$cand_nm)

# look at the first 100 values for disb_disc
cf$disb_desc[1:100]

# boxplots of expenditures by candidate
par(las=1, mar=c(3,12,1,1))
boxplot(cf$disb_amt ~ cf$cand_nm, horizontal = TRUE)
```

# Filter Data
This example uses the filter() function from the dplyr package, which is much easier to use than base R. Note that you can filter on multiple fields, or use conditions. Check the documentation for filter by typing '?filter' in the console.
```{r}
# load packages; if you haven't installed try running: install.packages('dplyr')
library(dplyr)

# filter to only include Trump and Clinton, filter to remove any disb_amt that is < 0; these must be data entry errors
cf_sub = filter(cf, cand_nm == 'Trump, Donald J.' | cand_nm == 'Clinton, Hillary Rodham', disb_amt >= 0)
```

#Task 1: Clean up text fields
```{r}
library(magrittr)
library(tidyr)
library(stringr)

# separate cand_nm into two columns; note that you need to choose how to handle cnadidates with > 2 names by specifying the 'extra' parameter.
# view the documentation for separate by typing ?separate to control the characters R will separate on (spaces, underscores, decimal, etc)
cf_clean = separate(cf_sub, col = cand_nm, into = c("cand_last", "cand_first"), extra = "merge")



# separate disb_dt into 3 columns for day, month, and year.
# before we separate, let's make a copy of disb_dt to use later when we explore DateTime objects.
cf_clean$disb_dt_str = cf_clean$disb_dt
cf_clean = separate(cf_clean, disb_dt, into = c("disb_day", "disb_month", "disb_yr"))
# there is also a function called unite() that is the opposite of separate--it combines 2 columns into one



# remove space OR comma + space from a text field using the regex ' |, '
# first, an example using x; then we actually perform this substitution on our dataframe
x = cf_clean$recipient_nm[1:10]
str_replace_all(x, ' |, ', '_')

cf_clean$recipient_nm = str_replace_all(cf_clean$recipient_nm, ' |, ', '_')
```

#More text cleaning: replacing values
```{r}

# use the recode function to revalue a variable (replacing values with new values)
# we'll view our values for disb_yr and then replace 15 and 16 with 2015 and 2016.
table(cf_clean$disb_yr)
x = cf_clean$disb_yr[1:50]
recode(x, `16` = "2016", `15` = "2015") # I have no idea why you need to use backticks for numeric values, but you do.

cf_clean$disb_yr = recode(cf_clean$disb_yr, `16` = "2016", `15` = "2015")
table(cf_clean$disb_yr)
cf_clean = filter(cf_clean, disb_yr == "2016" | disb_yr == "2015") # here, we drop the 3 pesky observations with a bad year



# same for months
table(cf_clean$disb_month)
x = cf_clean$disb_month[1:50]
recode(x, "JAN" = "January", "FEB" = "February", "MAR" = "March", "APR" = "April", "MAY" = "May", "JUN" = "June", "JUL" = "July", "AUG" = "August", "SEP" = "September", "OCT" = "October", "NOV" = "November", "DEC" = "December")

cf_clean$disb_month = recode(cf_clean$disb_month, "JAN" = "January", "FEB" = "February", "MAR" = "March", "APR" = "April", "MAY" = "May", "JUN" = "June", "JUL" = "July", "AUG" = "August", "SEP" = "September", "OCT" = "October", "NOV" = "November", "DEC" = "December")
table(cf_clean$disb_month)



# keep only first 5 digits of ZIP so all ZIP codes are the same
# we do this here by applying the substr() function to each element of the recipient_zip column
# note that you could apply ANY valud function to this column the same way using the mutate() function!
cf_clean = mutate(cf_clean, recipient_zip = substr(recipient_zip, 1, 5))
```

# Indicator variables, quantiles, and scaling
```{r}
library(Hmisc)
# binary indicator variables - the easy way
# create a TRUE/FALSE indicator for rent, based on the value in disb_disc column
cf_clean$is_rent = ifelse(cf_clean$disb_desc == "RENT", TRUE, FALSE)



# create indicator column that breakst disb_amt into quartiles; this is useful if you want to visualize/summarize within those groups
cf_clean$disb_amt_group = cut2(cf_clean$disb_amt, g = 4)



# create new column that is scaled disb_amt; this uses the default for scale() which is to scale to mean 0 standard deviation 1
# if you want to REPLACE the column instead of creating a new one, use: mutate(cf_clean, disb_amt = scale(disb_amt))
cf_clean = mutate(cf_clean, disb_amt_scaled = scale(disb_amt))
```

# Dates and times
```{r}
# convert date to date object; note that it LOOKS the same afterward but is a datetime object.
# to view format details, check ?strptime
x = cf_clean$disb_dt_str[0:10]
x[1]
as.Date(x, format = '%d-%b-%y')
cf_clean = mutate(cf_clean, disb_dt = as.Date(disb_dt_str, format = '%d-%b-%y'))



# create column with dat of week of date; note that this would only work on datetime objects (not on strings)
cf_clean = mutate(cf_clean, disb_weekday = weekdays(disb_dt))
```

# Anonymize/hash strings
**Disclaimer: I know nothing about hashing and you should ensure that whatever method you choose to hash your data is SECURE and LEGAL.**
```{r}
# hash cmte_id using sha256
cf_clean = mutate(cf_clean, cmte_id = sha256(as.character(cmte_id)))



# hash cand_id using md5
cf_clean = mutate(cf_clean, cand_id = md5(as.character(cand_id)))
```

# Drop columns
```{r}
# my preferred method: make a list; drop everything in that list
drop_cols = c("cmte_id", "cand_id", "cand_first") 
cf_clean = cf_clean[, !(names(cf_clean) %in% drop_cols)]
```

#Counts and other summaries by group
ddply() is your best friend here. If you need to do any kind of groupwise operation, this is probably the way to do it.

This is typically a useful way to get data for plots as well.

```{r}
# get counts by candidate; this is just number of records for each candidate
cand_counts = ddply(cf_clean, .(cand_last), nrow)
cand_counts

# get total disbursements by year and month
disb_yr_mo = ddply(cf_clean, .(disb_yr, disb_month), summarize, total_disb = sum(disb_amt))
disb_yr_mo

# get average disbursements by state
disb_st_avg = ddply(cf_clean, .(recipient_st), summarize, avg_disb = mean(disb_amt))
arrange(disb_st_avg, desc(avg_disb))
```

# Reshaping data
We'll use a different dataset here.
```{r}
library(tidyr)
data("occupationalStatus")
os = data.frame(occupationalStatus)
os

# convert from "narrow" to "wide" data by spreading destination column and using Freq as value
spread(os, destination, Freq)
```

# Basic exploratory data analysis
```{r}
barplot(table(cf_clean$cand_last))

# sums by date; here we aggregate first using the aggregate() function 
# aggregate() is a simpler, but more limited, tool to aggregate data by dimension
daily_sums = aggregate(disb_amt ~ disb_dt , data=cf_clean, sum)
plot(daily_sums, type = "l")

# what if we want to view this by part? the easiest way is to use a different package (see below); base R is quite limited in its plotting
```

# Exploratory data analysis & piping with ggplot2
Note: if you like this section and want to use more ggplot2, I would HIGLY recommend "The R Graphics Cookbook" by Winston Chang.

```{r}
library(ggplot2)

#
daily_sums_party = aggregate(disb_amt ~ disb_dt + cand_last, data=cf_clean, sum)
ggplot(daily_sums_party, aes(x = disb_dt, y = disb_amt, color = cand_last)) + geom_line()


# let's take our example from above (counts by candidate) and turn it into a simple bar graph
cf_clean %>%
    ddply(.(cand_last), nrow) %>%
    ggplot(aes(x = cand_last, y = V1)) + geom_bar(stat = "identity")

# more advanced: let's use the date (this is why we needed Date objects) to plot totals by date
cf_clean %>%
    ggplot(aes(x = disb_dt, y = disb_amt)) + geom_line() + facet_grid(cand_last ~ .)

# contributions by day of week
cf_clean %>%
    ddply(.(disb_weekday), summarize, total_disb = sum(disb_amt)) %>%
    ggplot(aes(x = disb_weekday, y = total_disb)) + geom_bar(stat = "identity")
    
```




